---
title: "DSI R Packages"
author: "Michael Culshaw-Maurer"
date: "1/19/2018"
output: html_document
---

# Basics of packages
- making R packages is easy, shouldn't be reluctant to do it
- what is a package?
    - it's an installable unit of R code
    - it's a single collection of things
    - it can be functions, it can be data, it can be both
    - it SHOULD have documentation, doesn't NEED to
    - it's easy to install and share with others
- incentives for organizing stuff into packages
    - much easier to install a package on another machine
        - ESPECIALLY when working on a cluster
    - really easy to send to someone else AND get help
        - **Duncan especially likes this when helping people**

# Working with packages

This is what you should use for using local packages
```{r, eval=F}
install.packages("package", repos=NULL)
install.packages("package.tar.gz", repos=NULL)
```

Use this to get stuff from github

- here, you're installing from source
- `install.packages` is probably downloading already binary
- you should install from CRAN when possible but it's often not (or not the most up-to-date)
```{r, eval=F}
devtools::install_github("dsidavis/RMinPackage")
```

Local packages from terminal command line
    - you can use a set of these commands for R in terminal, often fast and direct
```{bash, eval=F}
R CMD INSTALL package.tar.gz
R CMD INSTALL packageDirectory
```

# Where is the package installed

- library vs. package
    - a library is a directory which contains one or more installed packages
- R has recommended packages in its own R directory
    - when you install a new package, you typically don't have permission to write there
- R will ask you to create your own library directory
- you can create this **ahead of time** as any directory
- specify your own collection of R library directories
    - R_LIBS or R_LIBS_USER environment variable as a **colon-separated** list of directories
    - via `.libPaths()`
        - this w/o arguments tells you where it's looking, with an argument will change where it looks
- installing into personal "libraries"
- **should create one in the home directory called "R_Packages"**

`available.packages()` will go to CRAN mirror and give you lots of descriptions of packages
`listfiles(.libPaths()[1])` will list all the packages, then put all of these into `install.packages()` to reinstall them somewhere else

in your home directory you should have a ~/.RProfile where you can set `.libPaths()` and such

# Packages- Loading and Attaching
- Difference between loading a package and attaching to search path
- Loading: refer explicitly to a variable in a package, load another package that imports/depends on package
- Attaching: `library()` and `require()` will put that package in the search path so you can call functions
- doing something with double colons like `dplyr::filter()` will look outside of the normal search path
    - there's actually some computation going on when using the `::` but it's not bad
    - it can make your code more rigid so maybe shouldn't use it
- every package has its own little search path defined by its namespace
- `:::` will get out objects that aren't intentionally available (some packages have stuff in them that you're not really supposed to access). For example
- **the bigger the search path, the slower and more prone to error you'll be**
- double colons will LOAD the package, library() will load AND attach it (put it on the search path)
    - this is why you `detach()` packages
- apparently you **don't** want to use library() or require(), using double colons can be better

# R code in packages

- when writing a package for your own use, anything goes
    - don't need documentation, it's for you
- for others to use, you need:
    - the code to be robust to different inputs
        - check class/type of inputs, lengths, NAs
    - to provide meaningful error messages
    - to provide good documentation, examples, and vignettes
    - to answer questions promptly
    - to avoid reimplementing functions in existing packages
- make the functions flexible
    - add parameters to allow the caller to specify values you compute
    
# Why write a package?

- if you want to install on multiple machines
- very easy and gives you a single unit instead of loose files
- contributes to academic community
- **you can write a package and write a short paper about that package and get a pub out of it**

# Anatomy of a Package
- directory containing:
    - `DESCRIPTION` file
    - `NAMESPACE` file
    - `R` directory
- additional directories
    - `man` help files
    - `tests` tests to check the package
    - `data` data provided by the package
    - `inst` other files to be installed with the package
- Duncan has an RMinPackage on github "https://github.com/dsidavis/RPackagesWorkshop/tree/master/RMinPackage"
    - honestly just copy this minimal stuff, then make an R directory, copy all your R files into it, and there you go, you have a package
- `NAMESPACE` is where you define what you want to export to make available to people
    - only export the stuff you want people to be able to use, ignore any helper functions you have
- "Writing R extensions manual" is the bible of this stuff
- Hadley also has an "R Packages" book that's good

## Version Number

- `major-minor.patch`
- patch: fix a bug, documentation, etc. but no new features/functionality
- minor: new feature, like new function, new parameter for function
- major: 
    - cumulative critical mass of new functionality worthy of note
    - big change that makes it backwards-incompatible
    
## R Code

- Put R code in R directory
- typically a collection of functions
- code will be source()d at installation time

## Namespace

- will NOT look in search path, only on the things that you include as imports
- you could include several packages as imports, and the ORDER HERE MATTERS

## Keywords/Concepts

- underused, especially with roxygen
    - defined set of keywords you can use
- Concepts

## action when a package loaded

- when is it loaded but not attached?
    - when :: is used
    - when it's imported with another package
    
    
`R CMD check mypackage` will check everything about whether your package works, you should do this before you commit a file

`R CMD install --build` will build a binary version, which you can put on CRAN, but they'll probably ask you to change lots of stuff

**you can distribute yourself**
can also just use github (but don't wanna do this if you have proprietary data/code)