---
title: "TidyR and Joins"
author: "R.D. Avis"
date: "2/27/2018"
output: html_document
---

# Purpose
We are going to convert data from wide to long format and vice versa

First we need to load our libraries

```{r load_libraries, message = FALSE, warning = FALSE}
library(tidyverse)
```

Now let's get our data

```{r get_data}
surveys <- read_csv("https://gge-ucd.github.io/R-DAVIS/data/portal_data_joined.csv")

glimpse(surveys)
```

First, summarize these data.

```{r summarize}
surveys_gw <-
  surveys %>%
  filter(!is.na(weight)) %>% 
  group_by(genus, plot_id) %>% 
  summarize(mean_weight = mean(weight))

glimpse(surveys_gw)

head(surveys_gw)
```

Let's spread these data into wide format with a different plot id for each row, and different genera across columns

```{r spread}
surveys_gw %>% 
  spread(key = genus, value = mean_weight)

```

Specify the fill value

```{r fill_value_for_spread}
surveys_gw %>% 
  spread(key = genus, value = mean_weight, fill = 0)
```

Now we want to convert from wide format back to long format

```{r}
surveys_wide <-
  surveys_gw %>% 
  spread(key = genus, value = mean_weight, fill = 0)

surveys_wide

# When we gather(), we have to *think* about what the keys and values are

surveys_wide %>% 
  gather(key = genus, value = mean_weight, Baiomys:Spermophilus)

surveys_wide %>% 
  gather(key = the_genus, value = the_mean_weight, -plot_id)
```
# Challenge 1

Spread the `surveys` data frame with year as columns, `plot_id` as rows, and the number of genera per plot per year as the values. You will need to summarize before reshaping, and use the function `n_distinct()` to get the number of unique genera within a partiuclar chunk of data.

Hint: How do you need to summarize the data to count 




```{r}
genera_plot_year <- surveys %>%
  group_by(plot_id, year) %>% 
  summarise(num_genera = n_distinct(genus))

glimpse(genera_plot_year)

genera_plot_year %>% 
  spread(key = year, value = num_genera)
```


```{r make_a_csv}

surveys %>% 
  group_by(year, plot_id) %>% 
  summarize(unique_genera_count = n_distinct(genus)) %>% 
  spread(key = year, value = unique_genera_count) %>% 
  write_csv(path = "an.csv")

```

